\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% \noindent This article investigates the modelling of a simple channel 2d planar flow that is developed between flatplates to validate this flow a well known analytical solution from the literature is taken and further validated it with simulation results. This article shows the accuracy and sensitivity of the domain depending on the mesh sizes and imposed boundary conditions. As a continuous pde can never be inserted there are numerical errors that occur within the solution and quantification of these errors were discussed. Finally an isothermal simulation was ran to show the effect of temperature on the velocity of the flow.

\noindent This report investigates the fully developed incompressible 2D laminar flow in a radial pipe. An analytical velocity profile is derived from Navier-Stokes equation to validate with the simulations that were ran. This analytical equations assumes that the flow is fully developed, two dimensional and is in steady state. Further the analytical is discretised based on the number of grid points in the wall normal direction. The computational domain is increased according to the full developement of the flow and refined with three different mesh sizes. Two outlet boundary conditions were used to depict the changes in velocity profile at the outlet. These solutions were further compared with analytical solution with their respective grid points. The comparision between the two oulet conditions and analytical was performed with each mesh size. Comparision of the flow profile with changing the order of upwind scheme is assessed. The relative and absolute errors are presented for these comparisions at the outlet wall. A non isothermal simulation was performed with Dirichlet boundary condition at the top and bottom walls to study the affect of temperature on the simulation.
